Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-07-04 icfp24paper/paper.tex
update artifact, add data availability, add acknowledgement

Let us look at some basic examples. Recall the meta-level identity function; it
can be used at the object-level too, by applying it to quoted terms:
  \[ \letdef \vn : \Nat := \spl(\id\,\ql 10 + 10 \qr); ... \]
Here, $\id$ is used at type $\Lift \Nat$. During unstaging, the expression in
the splice is evaluated, so we get $\spl\ql 10 + 10 \qr$, which is
definitionally the same as $10 + 10$, which is our code output here. Boolean
short-circuiting is another basic use-case:
\begin{alignat*}{3}
  &\mit{and} : \Up\Bool \to \Up\Bool \to \Up\Bool\\
  &\mit{and}\,\vx\,\vy = \ql\case \spl \vx\,\of \True \to \spl \vy; \False \to \False\qr
\end{alignat*}
Since the $y$ expression is inlined under a $\case$ branch at every use site, it
is only computed at runtime when $x$ evaluates to $\True$. In many situations,
staging can be used instead of laziness to implement short-circuiting, and often
with better runtime performance, avoiding the overhead of thunking. Consider the
$\map$ function now:
\begin{alignat*}{3}
  & \hspace{-5em}\mathrlap{\map : \{\vA\,\vB : \VTy\} \to (\Up \vA \to \Up \vB) \to \Up (\List\,\vA) \to \Up(\List\,\vB)}\\
  & \hspace{-5em}\mathrlap{\map\,\vf\,\vas = \ql\letrec \go\,\vas := \case \vas\,\of}\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Nil             &&\to \Nil;\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Cons\,\va\,\vas &&\to \Cons\,\spl(\vf\,\ql \va \qr)\,(\go\,\vas);\\
  & \hspace{-5em}\ind\ind\ind\ind\ind\go\,\spl\vas \qr
\end{alignat*}
For example, this can be used as $\letdef \vf\,\vas : \List\,\Nat \to
\List\,\Nat := \spl(\map\,(\lam \vx.\,\ql \spl \vx + 10 \qr)\,\qt{\vas}) $.
This is unstaged to a recursive definition where the mapping function is inlined
into the $\Cons$ case as $\Cons\,\va\,\vas \to \Cons\,(\va + 10)\,(\go\,\vas)$.
Note that $\map$ has to abstract over value types, since lists can only contain
values, not functions. Also, the mapping function has type $\Up \vA \to \Up
\vB$, instead of $\Up (\vA \to \vB)$. The former type is often preferable to the
latter; the former is a metafunction with useful computational content, while
the latter is merely a black box that computes object code. If we have $\vf :
\Up(\vA \to \vB)$, and $\vf$ computes to $\ql \lam \vx.\,\vt\qr$, then $\spl
\vf\,\vu$ is unstaged to an undesirable ``administrative'' $\beta$-redex $(\lam
\vx.\,\vt)\,\vu$.

@@

@@

Unstaging is defined as evaluation of the syntax of CFTT in a particular
presheaf model.
\begin{itemize}
  \item The \emph{object theory} is the simple type theory supporting only value
    types and computation types as described in Section
    \ref{sec:the-object-level} and also in the Agda
    formalization. Object-theoretic terms are not quotiented by
    $\beta\eta$-conversion. Substitution is a recursively defined operation on
    terms.
  \item $\mbbo$ is the category whose objects are contexts of the object theory,
    and morphisms are parallel substitutions, as lists of terms. With this,
    object-theoretic types form a constant presheaf which we denote
    as $\Tyo : \mbbo^{\msf{op}} \to \Set$, and terms form a dependent presheaf over $\Tyo$ which we write as
    $\Tmo$.
  \item
  In the presheaf model of CFTT, every typing context (and also every closed
  type) is modeled as a presheaf over $\mbbo$. The interpretation of meta-level
  type formers is standard, following Hofmann \cite{Hofmann97syntaxand} and
  Huber \cite{huber-thesis}.
\item $\Ty : \MTy_0\,\Gamma$ is the displayed presheaf which is constantly
  $\Tyo$. $\VTy$ and $\CTy$ are modeled analogously.
\item $\Lift A : \MTy_0\,\Gamma$ is modeled as follows. We have that $A :
  \MTy\,\Gamma\,\Ty$, but since $\Ty$ is defined as constantly $\Tyo$, it
  follows that $A$ is a natural transformation from $\Gamma$ to $\Tyo$. Hence,
  we define $\Lift A : \MTy_0\,\Gamma$ to be $\Tmo[A]$, where $[\blank]$ is
  substitution in presheaves, i.e.\ the reindexing of a displayed presheaf by a
  natural transformation.
\item
  $\Tm\,\Gamma\,A$ is simply defined to be $\MTy\,\Gamma\,(\Lift A)$ in the model. Hence,
  quotation and splicing are both interpreted as identity maps.
\end{itemize}
Now, we can take a closed term $t : \Tm\,\emptycon\,A$ in CFTT and evaluate it
in the presheaf model to obtain an object-theoretic term whose type is the
result of evaluating $A$.

>>>>

Let us look at some basic examples. Recall the meta-level identity function; it
can be used at the object-level too, by applying it to quoted terms:
  \[ \letdef \vn : \Nat := \spl(\id\,\ql 10 + 10 \qr); ... \]
Here, $\id$ is used at type $\Lift \Nat$. During unstaging, the expression in
the splice is evaluated, so we get $\spl\ql 10 + 10 \qr$, which is
definitionally the same as $10 + 10$, which is our code output here. Boolean
short-circuiting is another basic use-case:
\begin{alignat*}{3}
  &\mit{and} : \Up\Bool \to \Up\Bool \to \Up\Bool\\
  &\mit{and}\,\vx\,\vy = \ql\case \spl \vx\,\of \True \to \spl \vy; \False \to \False\qr
\end{alignat*}
Since the $y$ expression is inlined under a $\mbf{case}$ branch at every use site, it
is only computed at runtime when $x$ evaluates to $\True$. In many situations,
staging can be used instead of laziness to implement short-circuiting, and often
with better runtime performance, avoiding the overhead of thunking. Consider the
$\map$ function now:
\begin{alignat*}{3}
  & \hspace{-5em}\mathrlap{\map : \{\vA\,\vB : \VTy\} \to (\Up \vA \to \Up \vB) \to \Up (\List\,\vA) \to \Up(\List\,\vB)}\\
  & \hspace{-5em}\mathrlap{\map\,\vf\,\vas = \ql\letrec \go\,\vas := \case \vas\,\of}\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Nil             &&\to \Nil;\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Cons\,\va\,\vas &&\to \Cons\,\spl(\vf\,\ql \va \qr)\,(\go\,\vas);\\
  & \hspace{-5em}\ind\ind\ind\ind\ind\go\,\spl\vas \qr
\end{alignat*}
For example, this can be used as $\letdef \vf\,\vas : \List\,\Nat \to
\List\,\Nat := \spl(\map\,(\lam \vx.\,\ql \spl \vx + 10 \qr)\,\qt{\vas}) $.
This is unstaged to a recursive definition where the mapping function is inlined
into the $\Cons$ case as $\Cons\,\va\,\vas \to \Cons\,(\va + 10)\,(\go\,\vas)$.
Note that $\map$ has to abstract over value types, since lists can only contain
values, not functions. Also, the mapping function has type $\Up \vA \to \Up
\vB$, instead of $\Up (\vA \to \vB)$. The former type is often preferable to the
latter; the former is a metafunction with useful computational content, while
the latter is merely a black box that computes object code. If we have $\vf :
\Up(\vA \to \vB)$, and $\vf$ computes to $\ql \lam \vx.\,\vt\qr$, then $\spl
\vf\,\vu$ is unstaged to an undesirable ``administrative'' $\beta$-redex $(\lam
\vx.\,\vt)\,\vu$.

@@

\section*{Data Availability Statement}
The Agda and Haskell artifacts are available in \cite{cftt-artifacts}.

@@

Unstaging is defined as evaluation of the syntax of CFTT in a particular
presheaf model.
\begin{itemize}
  \item The \emph{object theory} is the simple type theory supporting only value
    types and computation types as described in Section
    \ref{sec:the-object-level} and also in the Agda
    formalization. Object-theoretic terms are not quotiented by
    $\beta\eta$-conversion. Substitution is a recursively defined operation on
    terms.
  \item $\mbbo$ is the category whose objects are contexts of the object theory,
    and morphisms are parallel substitutions, as lists of terms. With this,
    object-theoretic types form a constant presheaf which we denote
    as $\Tyo : \mbbo^{\msf{op}} \to \Set$, and terms form a dependent presheaf over $\Tyo$ which we write as
    $\Tmo$.
  \item
  In the presheaf model of CFTT, every typing context (and also every closed
  type) is modeled as a presheaf over $\mbbo$. The interpretation of meta-level
  type formers is standard, following \cite{Hofmann97syntaxand} and \cite{huber-thesis}.
\item $\Ty : \MTy_0\,\Gamma$ is the displayed presheaf which is constantly
  $\Tyo$. $\VTy$ and $\CTy$ are modeled analogously.
\item $\Lift A : \MTy_0\,\Gamma$ is modeled as follows. We have that $A :
  \MTy\,\Gamma\,\Ty$, but since $\Ty$ is defined as constantly $\Tyo$, it
  follows that $A$ is a natural transformation from $\Gamma$ to $\Tyo$. Hence,
  we define $\Lift A : \MTy_0\,\Gamma$ to be $\Tmo[A]$, where $[\blank]$ is
  substitution in presheaves, i.e.\ the reindexing of a displayed presheaf by a
  natural transformation.
\item
  $\Tm\,\Gamma\,A$ is simply defined to be $\MTy\,\Gamma\,(\Lift A)$ in the model. Hence,
  quotation and splicing are both interpreted as identity maps.
\end{itemize}
Now, we can take a closed term $t : \Tm\,\emptycon\,A$ in CFTT and evaluate it
in the presheaf model to obtain an object-theoretic term whose type is the
result of evaluating $A$.
